--- 
title: "Delivering Data Science Products via Packages"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: bradleyboehmke/misk-packages
description: "Creating R and Python packages to scale data science."
---

# Welcome

Packages are the fundamental units of reproducible R and Python code. They include reusable components, the documentation that describes how to use them, requirements to ensure the user can apply them and tests to ensure consistent and reliable functionality. In this book you’ll learn how to turn your code into packages so that you, your teammates, and others can easily download and use. Writing a package can seem overwhelming at first. So start with the basics and improve it over time. It doesn’t matter if your first version isn’t perfect as long as the next version is better.

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```

## Who should read this {-}

```{block, type='todo'}
TBD
```

This module makes a few assumptions of your established knowledge regarding your programming skills. Below are some important assumptions made and resources to read through if you feel inadequately prepared:

```{block, type='todo'}
Complete after rest of material is finalized.
```

| Assumptions                       | `r fontawesome::fa("python")` Resource | `r fontawesome::fa("r-project")` Resource   
| --------------------------------- | ------------- | ------------- |
| You should be familiar with...    |               |               | 
| You should be familiar with...    |               |               |
| You should be familiar with...    |               |               |
| You should be familiar with...    |               |               |

## What you need for this book {-}

```{block, type='todo'}
TBD
```

## Conventions used in this book {-}

The following typographical conventions are used in this book:

* ___strong italic___: indicates new terms,
* __bold__: indicates package & file names,
* `inline code`: monospaced highlighted text indicates functions or other commands that could be typed literally by the user,
* code chunk: indicates commands or other text that could be typed literally by the user

```{r, first-code-chunk, collapse=TRUE}
1 + 2
```

We will also refer to both R and Python throughout. When discussing one or the other we will typically label with the appropriate icons:

- Python: `r fontawesome::fa("python")`
- R: `r fontawesome::fa("r-project")`

In addition to the general text used throughout, you will notice the following code chunks:

```{block, type = "tip"}
Signifies a tip or suggestion
```

```{block, type = "note"}
Signifies a general note
```

```{block, type = "warning"}
Signifies a warning or caution
```

## Feedback {-}

Reader comments are greatly appreciated. To report errors or bugs please post an issue at https://github.com/misk-data-science/misk-packages/issues.

## Acknowledgments {-} 

```{block, type='todo'}
TBD
```
